---
title: "R Basics: Data Analysis"
format: html
editor: 
  markdown: 
    wrap: 72
---

```{r include=FALSE}
knitr::purl("basic-data-analysis.qmd")
```


## Data ingestion

The first step when doing any kind of data analysis with code is getting
the data in and cleaning it up. Compared to using Excel, GraphPad, etc,
this is arguably the point of highest relative friction when working
with code. Programming languages prefer data in a neat and tidy tabular
format, and usually it isn't like that to begin with, so we have to do
some cleaning and reshaping.

This takes some work, but depending on the purposes, is often worth it
in the end.

```{r warning=FALSE, message=FALSE}
#| column: page

library(readxl) # A useful library for reading Excel files

alamar_raw_plate <- read_excel(
  "jt58 etoposide kill curve 24h post eto wt vs 3ko13 3ko17 alamar.xlsx",
  range = "B15:M22", # For plate data, know where your plate reader puts it!
  col_names = FALSE
)

# Or, since we only use read_excel, we could just write `readxl::read_excel`
# (This is better practice than loading the entire library for one function)

alamar_raw <- alamar_raw_plate[1:7, 2:10] # Trim away empty wells
# Or, equivalently: alamar_raw <- alamar_raw_plate[-8, 2:10]

columns <- c(
  "WT_Par_1", "WT_Par_2", "WT_Par_3",
  "M3KO_13_1", "M3KO_13_2", "M3KO_13_3",
  "M3KO_17_1", "M3KO_17_2", "M3KO_17_3"
)

colnames(alamar_raw) <- columns

# Get and subtract the control wells

background <- as.numeric(alamar_raw[1,]) # `mean()` won't work on a data frame
avg_background <- mean(background)

alamar_corrected <- alamar_raw - avg_background # Subtracts from each well
alamar_corrected <- alamar_corrected[-1,] # Remove the control wells

# Set the concentrations and add to the table

etoposide_concs <- rev(c(0, 10^seq(0, 2, by = 0.5)))
alamar_corrected$conc <- etoposide_concs

# Print out a preview of the data frame
knitr::kable(alamar_corrected) # kable() gives us the table in a pretty format
```

## Reshaping the data

The data is still not in the format modelling and plotting functions
expect.

The key to tidy data is simple enough: one column per variable. This is
the format that most data analysis software expects or tries to force
data to be in.

The "why" of tidy data is discussed at length by Hadley Wickham, chief
scientist at Posit (who make RStudio). If interested, take a gander:
https://vita.had.co.nz/papers/tidy-data.pdf

For a more practical guide, you can look here:
https://r4ds.had.co.nz/tidy-data.html

In this code, I use the `tidyr` library, and there is a *very* helpful
cheetsheet that I basically always check when I need to use it, if you
just want a quick answer on how to do a given thing:
https://rstudio.github.io/cheatsheets/tidyr.pdf

The `pivot_longer` function here lets us turn the unique columns for
each genotype and replicate in the above into a simpler data frame with
one genotype column and one replicate column. In practise, we don't need
the replicate column but I've included it here just to be explicit.

```{r warning=FALSE, message=FALSE}

alamar <- tidyr::pivot_longer(
  data = alamar_corrected,
  cols = -conc, # all columns except `conc`
  names_sep = "_",
  names_to = c("genotype", "clone", "replicate"),
  values_to = "corrected_absorbance"
)

knitr::kable(head(alamar, 10)) # Just show the first 10 rows
```

## Modelling and plotting

Now that we have the data in the right format, we can use R's tools for
making statistical models and plotting them. Since we have dose-response
data, we can use the `drc` package. Normally, dose-response curves in
biology are not linear, but S-shaped, and we typically model S-shaped
data with so called "logistic" models, usually the "4-parameter"
logistic model (the 4 parameters are: minimum, maximum, slope steepness,
and inflection point). `drc` gives us easy access to a whole family of
dose response models, so we don't really have to get into the maths.

```{r warning=FALSE, message=FALSE}
library(drc) # Both drm & LL.4() are from this library, so OK to import it all

dose_response_model <- drm(
  corrected_absorbance ~ conc, # Model absorbance as a function of concentration
  fct = LL.4(), # This says to use the 4-parameter logistic model
  curveid = genotype, # This says: do a separate model/curve for each genotype
  data = alamar # specifies the dataframe
)

ec50 <- ED(dose_response_model, 50) # (We can choose any %, not just 50%)

plot(dose_response_model)
abline(v=c(ec50[1],ec50[2]), col=c("blue", "red"), lty=c(1,3))
```

## Making the plot prettier

The base R plots are fine for a quick and dirty inspection, but
depending on what you want to show, there are far more powerful
libraries.

```{r warning=FALSE, message=FALSE}
library(ggplot2)
library(ggprism)

# First set the genotype to a factor so we can control the order (for legends)
# (ggplot will convert it under the hood, and the default order is alphabetical)
alamar$genotype <- factor(alamar$genotype, levels = c("WT", "M3KO"))

# Then we make the labels useful for plotting
levels(alamar$genotype) <- c("Wild-type", "IFITM3 KO")

max_y <- max(alamar$corrected_absorbance)

plot <- ggplot(
  alamar,
  aes(x = conc, y = corrected_absorbance, color = genotype, fill = genotype)
)

plot <- plot +
  geom_line(
    stat = "smooth",
    method = drm,
    method.args = list(fct = LL.4()),
    linewidth = 2,
    se = FALSE
  ) +
  geom_point(
    color = "black",
    pch = 21,
    size = 3,
    alpha = 0.8,
    stroke = 1.2
  ) +
  scale_color_manual(values = c("#5278ec", "#bb0c40")) +
  scale_fill_manual(values = c("#5278ec", "#bb0c40")) +
  theme_prism(base_size = 14) +
  scale_y_continuous(limits = c(NA, max_y * 1.1)) +
  scale_x_continuous(
    trans = scales::pseudo_log_trans(0.1),
    labels = scales::label_number(drop0trailing = TRUE),
    breaks = c(0, 0.3, 1, 3, 10, 30, 100)
  ) +
  labs(x = "Etoposide (µM)", y = "Abs (450)")

# add the EC50 lines
plot <- plot +
  geom_vline(
    xintercept = ec50[1],
    color = "#5278ec",
    linewidth = 1.1,
    linetype = 2
  ) +
  geom_vline(
    xintercept = ec50[2],
    color = "#bb0c40",
    linewidth = 1.1,
    linetype = 3
  )

plot$theme[c("legend.text.align", "legend.title.align")] <- NULL

plot
```
